Polynomial Time and Dependent Types

### 2023-10-25 paper/paper.tex
[ paper ] more minor fixes, fix binomials, and tidy rules

Typical Type Theory only speaks of the public face that programs
present to the world -- if you input things like this, you get things
that look like that -- but cannot bring itself to mention the true
cost of programs' execution. One can encode costs by embedding another
programming language in Type Theory, for example \cite{GueneauCP18},
or one can synthesise costs by treating resource counting as a
computational effect, for example \cite{Danielsson08,NiuSGH22}, but
neither of these capture the intrinsic costs of the programs we write
in Type Theory. These techniques deliver only \emph{conspicuous
  consumption}, not speaking of the real resources consumed.

@@

For this section, the affine linear $\lambda$-calculus we will use
will have linear functions and $\otimes$-products. Contexts are
treated up to permutation, so uses of exchange are implicit.
\begin{mathpar}
  \inferrule*
  { }
  {\Gamma, x : A \vdash x : A}

@@

  \inferrule*
  {\Gamma_1 \vdash M : \Diamond \\ \Gamma_2 \vdash N : \tyNat}
  {\Gamma_1, \Gamma_2 \vdash \conSucc(M, N) : \tyNat}
\end{mathpar}
To construct a $\conZero$, we must have a $\Diamond$ to pay for it,
and likewise, to construct a $\conSucc$ we must pay a $\Diamond$. We
can think of $\Diamond$s as an unit of iterable data. Iterability is
``saved up'' in data during construction, and released during
iteration. Diamonds cannot be created by a program itself, for the
same reason that constructors were prohibited in the \emph{Cons-free}
system, but they are released from iterable data during iteration. The
LFPL natural number iterator has the following typing rule:
\begin{displaymath}
  \inferrule*
  {d : \Diamond \vdash M_z : A \\ d : \Diamond, x : A \vdash M_s : A \\ \Gamma \vdash N : \tyNat}
  {\Gamma \vdash \tmRec\,N\,\{\conZero(d) \mapsto M_z; \conSucc(d,x) \mapsto M_s \} : A}
\end{displaymath}
The difference with the \emph{Cons-free} iterator above is that the
$\conZero$ and $\conSucc$ cases now both have an additional binding of
type $\Diamond$. This allows some form nesting of iterations: during
an iteration over the input, the program can accumulate $\Diamond$s to
use for iteration over substructures that are smaller than the current
point in the iteration. A construction, due to \citet{AehligS02},
illustrates how this leads to all polytime computations. As above, we
assume a polynomial $p(n)$ and a step function
$f : \mathit{St} \multimap \mathit{St}$ that needs to be iterated
$p(n)$ times. We construct a linear iterator:
\begin{displaymath}
  \begin{array}{l}
    I_1 : (\tyNat \otimes \mathit{St}) \multimap (\tyNat \otimes \mathit{St}) \\
    I_1 = \lambda (n, s).\,\tmRec\,n\,\{
    \begin{array}[t]{@{}lcl}
      \conZero(d)&\mapsto&(\conZero(d), s);\\
      \conSucc(d,(n,s)) &\mapsto& (\conSucc(d,n),f\,s) \}
    \end{array}
  \end{array}
\end{displaymath}
Note that this iterator returns the natural number input as well as
the new state. LFPL does not allow duplication of iterable inputs, so
we must always reconstruct it if we want to do further
iteration. Addition of polynomials is accomplished by composition of
iterators. To raise the degree, we again use a nesting iterator:
\begin{displaymath}
  \begin{array}{l}
    I_{k+1} : (\tyNat \otimes \mathit{St}) \multimap (\tyNat \otimes \mathit{St}) \\
    I_{k+1} = \lambda (n,s).\,\tmRec\,n\,\{
    \begin{array}[t]{@{}lcl}
      \conZero(d)&\mapsto&(\conZero(d),s); \\
      \conSucc(d,(n,s)) &\mapsto& \Let\,(n,s)=I_k\,(n,s)\,\In\,(\conSucc(d,n), f\,s) \}
    \end{array}
  \end{array}
\end{displaymath}
Unlike in the \emph{Cons-free} system, this iterator does not raise
the degree of the nested iterator directly. Rather, the iterator $I_k$
on the input $n$ performs $\binom{n}{k}$ iterations. As observed by
Aelhig and Schwictenberg, this is sufficient because the binomials
form a basis for the vector space of all polynomials.

@@

In this section, we give an overview of the rules of QTT.

The full rules are presented in \autoref{sec:rules}.

The full rules are presented in the online appendix at

@@

The type $\mathrm{ND}$ supports a monad interface via the usual free
monad construction, as well as an effect
$\mathrm{flip} \stackrel1: \mathrm{ND}\,\BoolTy$ providing access to a
bit of non-deterministic information. Thus a program of type
$A \to \mathrm{ND}\,B$ in the $\sigma = 1$ fragment will be a polytime
program with access to an oracle. In the $\sigma = 0$ fragment, we can
write a function that resolves non-determinism using a list of
booleans. This function returns $\mathsf{nothing}$ if the list of
booleans is insufficient to resolve all the $\mathsf{choice}$s:
\begin{displaymath}
  \mathrm{runWithOracle} \stackrel0: \mathrm{ND}\,A \to \ListTy(\BoolTy) \to \mathrm{Maybe}\,A
\end{displaymath}
With these definitions, we can define Non-deterministic Polynomial
time as a predicate on problems:
\begin{displaymath}
  \mathrm{NP}(A,P) =
  \begin{array}[t]{@{}l}
    (f \stackrel1: \mathbf{R}(A \to \mathrm{ND}(\BoolTy)))\, \otimes \\
    \quad \left((a \stackrel1: A) \to \left((\mathit{bs} \stackrel1: \ListTy(\BoolTy)) \otimes (\mathrm{runWithOracle}\,(\mathbf{R}^{-1}(f)\,a)\,\mathit{bs} = \mathsf{just}\,\cTrue)\right) \Leftrightarrow P\,a\right)
  \end{array}
\end{displaymath}
Thus, a problem is in NP if there is a non-deterministic
boolean-valued polynomial time function that has a path to returning
$\cTrue$ exactly when the input satisfies the predicate. Moreover, it is a
quick matter of programming to see that problems in NP are closed
under the type of polytime reductions given above.

@@

  \ldots

  \inferrule* [right=Tm-U-Eq]
  {\Gamma \vdash M \stackrel\sigma: \TySet \\
    \Gamma \vdash N_1 \stackrel\sigma: \TyEl(M) \\
    \Gamma \vdash N_2 \stackrel\sigma: \TyEl(M)}
  {\Gamma \vdash N_1 =_M N_2 \stackrel\sigma: \TySet}
\end{mathpar}
Elimination:
\begin{displaymath}
  \inferrule* [right=Ty-El]
  {0\Gamma \vdash M \stackrel0: \TySet}
  {0\Gamma \vdash \TyEl(M)~\istype}
\end{displaymath}
Equality:
\begin{mathpar}
  \inferrule* [right=Ty-Eq-El-Pi]
  {0\Gamma \vdash M \stackrel0: \TySet \\
    0\Gamma, x \stackrel0: \TyEl(M) \vdash N \stackrel0: \TySet}
  {0\Gamma \vdash \TyEl((x \stackrel\rho: M) \to N) \equiv (x \stackrel\rho: \TyEl(M)) \to \TyEl(N)}

  \inferrule* [right=Ty-Eq-El-Bool]
  {0\Gamma~\isctxt}
  {0\Gamma \vdash \TyEl(\BoolTy) \equiv \BoolTy}

  \ldots

  \inferrule* [right=Tm-Eq-Set-Pi-Cong]
  {0\Gamma \vdash M \equiv M' \stackrel0: \TySet \\
    0\Gamma, x \stackrel0: \TyEl(M) \vdash N \equiv N' \stackrel0: \TySet}
  {0\Gamma \vdash (x \stackrel\pi: M) \to N \equiv (x \stackrel\pi: M') \to N' \stackrel0: \TySet}
\end{mathpar}

>>>>

Typically, Type Theory only speaks of the public face that programs
present to the world -- if you input things like this, you get things
that look like that -- but cannot bring itself to mention the true
cost of programs' execution. One can encode costs by embedding another
programming language in Type Theory, for example \cite{GueneauCP18},
or one can synthesise costs by treating resource counting as a
computational effect, for example \cite{Danielsson08,NiuSGH22}, but
neither of these capture the intrinsic costs of the programs we write
in Type Theory. These techniques deliver only \emph{conspicuous
  consumption}, not speaking of the real resources consumed.

@@

For this section, the affine linear $\lambda$-calculus we will use
will have linear functions and $\otimes$-products. Contexts are
treated up to permutation of entries, so uses of exchange are
implicit.
\begin{mathpar}
  \inferrule*
  { }
  {\Gamma, x : A \vdash x : A}

@@

  \inferrule*
  {\Gamma_1 \vdash M : \Diamond \\ \Gamma_2 \vdash N : \tyNat}
  {\Gamma_1, \Gamma_2 \vdash \conSucc(M, N) : \tyNat}
\end{mathpar}
To construct a $\conZero$, we must have a $\Diamond$ to pay for it,
and likewise, to construct a $\conSucc$ we must pay a $\Diamond$. We
can think of $\Diamond$s as an unit of iterable data. Iterability is
``saved up'' in data during construction, and released during
iteration. Diamonds cannot be created by a program itself, for the
same reason that constructors were prohibited in the \emph{Cons-free}
system, but they are released from iterable data during iteration. The
LFPL natural number iterator has the following typing rule:
\begin{displaymath}
  \inferrule*
  {d : \Diamond \vdash M_z : A \\ d : \Diamond, x : A \vdash M_s : A \\ \Gamma \vdash N : \tyNat}
  {\Gamma \vdash \tmRec\,N\,\{\conZero(d) \mapsto M_z; \conSucc(d,x) \mapsto M_s \} : A}
\end{displaymath}
The difference with the \emph{Cons-free} iterator above is that the
$\conZero$ and $\conSucc$ cases now both have an additional binding of
type $\Diamond$. This allows some form nesting of iterations: during
an iteration over the input, the program can accumulate $\Diamond$s to
use for iteration over substructures that are smaller than the current
point in the iteration. A construction, due to \citet{AehligS02},
illustrates how this leads to all polytime computations. As above, we
assume a polynomial $p(n)$ and a step function
$f : \mathit{St} \multimap \mathit{St}$ that needs to be iterated
$p(n)$ times. We construct a linear iterator:
\begin{displaymath}
  \begin{array}{l}
    I_1 : (\tyNat \otimes \mathit{St}) \multimap (\tyNat \otimes \mathit{St}) \\
    I_1 = \lambda (n, s).\,\tmRec\,n\,\{
    \begin{array}[t]{@{}lcl}
      \conZero(d)&\mapsto&(\conZero(d), s);\\
      \conSucc(d,(n,s)) &\mapsto& (\conSucc(d,n),f\,s) \}
    \end{array}
  \end{array}
\end{displaymath}
Note that this iterator returns the natural number input as well as
the new state. LFPL does not allow duplication of iterable inputs, so
we must always reconstruct it if we want to do further
iteration. Addition of polynomials is accomplished by composition of
iterators. To raise the degree, we again use a nesting iterator:
\begin{displaymath}
  \begin{array}{l}
    I_{k+1} : (\tyNat \otimes \mathit{St}) \multimap (\tyNat \otimes \mathit{St}) \\
    I_{k+1} = \lambda (n,s).\,\tmRec\,n\,\{
    \begin{array}[t]{@{}lcl}
      \conZero(d)&\mapsto&(\conZero(d),s); \\
      \conSucc(d,(n,s)) &\mapsto& \Let\,(n,s)=I_k\,(n,s)\,\In\,(\conSucc(d,n), s) \}
    \end{array}
  \end{array}
\end{displaymath}
Unlike in the \emph{Cons-free} system, this iterator does not raise
the degree of the nested iterator directly. Rather, the iterator $I_k$
on the input $n$ performs $\binom{n}{k}$ iterations. As observed by
Aelhig and Schwictenberg, this is sufficient because the binomials
form a basis for the vector space of all polynomials.

@@

In this section, we give an overview of the rules of QTT. The full
rules are presented in \autoref{sec:rules}.

\bob{The full rules are presented in the online appendix at}

@@

The type $\mathrm{ND}$ supports a monad interface via the usual free
monad construction, as well as an effect
$\mathrm{flip} \stackrel1: \mathrm{ND}\,\BoolTy$ providing access to a
bit of non-deterministic information. Thus a program of type
$A \to \mathrm{ND}\,B$ in the $\sigma = 1$ fragment will be a polytime
program with access to an oracle. In the $\sigma = 0$ fragment, we can
write a function that resolves non-determinism using a list of
booleans. This function returns $\mathsf{nothing}$ if the list of
booleans is insufficient to resolve all the $\mathsf{choice}$s:
\begin{displaymath}
  \mathrm{runWithOracle} \stackrel0: \mathrm{ND}\,A \to \ListTy(\BoolTy) \to \mathrm{Maybe}\,A
\end{displaymath}
With these definitions, we can define Non-deterministic Polynomial
time as a predicate on problems:
\begin{displaymath}
  \mathrm{NP}(A,P) =
  \begin{array}[t]{@{}l}
    (f \stackrel1: \mathbf{R}(A \to \mathrm{ND}(\BoolTy)))\, \otimes \\
    \, \left((a \stackrel1: A) \to \left((\mathit{bs} \stackrel1: \ListTy(\BoolTy)) \otimes (\mathrm{runWithOracle}\,(\mathbf{R}^{-1}(f)\,a)\,\mathit{bs} = \mathsf{just}\,\cTrue)\right) \Leftrightarrow P\,a\right)
  \end{array}
\end{displaymath}
Thus, a problem is in NP if there is a non-deterministic
boolean-valued polynomial time function that has a path to returning
$\cTrue$ exactly when the input satisfies the predicate. Moreover, it is a
quick matter of programming to see that problems in NP are closed
under the type of polytime reductions given above.

@@

  \ldots
\end{mathpar}
Elimination:
\begin{displaymath}
  \inferrule* [right=Ty-El]
  {0\Gamma \vdash M \stackrel0: \TySet}
  {0\Gamma \vdash \TyEl(M)~\istype}
\end{displaymath}
Equality:
\begin{mathpar}
  % \inferrule* [right=Ty-Eq-El-Pi]
  % {0\Gamma \vdash M \stackrel0: \TySet \\
  %   0\Gamma, x \stackrel0: \TyEl(M) \vdash N \stackrel0: \TySet}
  % {0\Gamma \vdash \TyEl((x \stackrel\rho: M) \to N) \equiv (x \stackrel\rho: \TyEl(M)) \to \TyEl(N)~\istype}

  % \inferrule* [right=Ty-Eq-El-Bool]
  % {0\Gamma~\isctxt}
  % {0\Gamma \vdash \TyEl(\BoolTy) \equiv \BoolTy~\istype}

  % \ldots

  \inferrule* [right=Ty-Eq-El-Cong]
  {0\Gamma \vdash M \equiv N \stackrel0: \TySet}
  {0\Gamma \vdash \TyEl(M) \equiv \TyEl(N)~\istype}
\end{mathpar}
